---
title: "Data Visualisation for Data Science"
author: "Hicham Zmarrou"
date: "Notebook -- <http://bit.ly/2pHMYck>  <br /> <br />"
output:
  html_document:
    toc: yes
  html_notebook:
    highlight: pygments
    number_sections: no
    theme: cosmo
    toc: yes
    toc_float: yes
subtitle: Introduction to dynamic graphics with rCharts
venue: TNO data science courses
---


<hr>

[Visit my website](http://trefoil.ml/) for more like this!

__References__

Most of this material is borrowed from:

* [rCharts website](https://ramnathv.github.io/rCharts/); and

* [dygraphs website](https://rstudio.github.io/dygraphs/index.html)
______________________________________________________________________________________________________________________________________

  


# rCharts

rCharts is an R package to create, customize and publish interactive javascript visualizations from R using a familiar lattice style plotting interface.

Before we begin charting, let us customize some settings for rCharts and knitr

```{r}
library(rCharts)
library(plyr)
library(knitr)
library(reshape2)
library(scales)
options(RCHART_WIDTH = 600, RCHART_HEIGHT = 400)
knitr::opts_chunk$set(comment = NA, results = 'asis', tidy = F, message = F)

```

## Polychart

We will create our first chart using Polychart, a javascript charting library based on the grammar of graphics, and inspired by ggplot2.

### Facetted Scatterplot

```{r, echo=TRUE}
names(iris) = gsub("\\.", "", names(iris))
p1 = rPlot(SepalLength ~ SepalWidth | Species, data = iris, color = 'Species', type = 'point')
p1$print(include_assets=T)

```

### Facetted Barplot

```{r}
hair_eye = as.data.frame(HairEyeColor)
p2 = rPlot(Freq ~ Hair, color = 'Eye', data = hair_eye, type = 'bar')
p2$facet(var = 'Eye', type = 'wrap', rows = 2)
p2$print(include_assets=T)
```

### Boxplot

```{r}
data(tips, package = 'reshape2')
p3 = rPlot(x = 'day', y = 'box(tip)', data = tips, type = 'box')
p3$print(include_assets=T)

```

### Column Chart

```{r}
dat = count(mtcars, .(gear, am))
p4 = rPlot(x = 'bin(gear, 1)', y = 'freq', data = dat, type = 'bar', 
  list(var = 'am', type = 'wrap'))
p4$print(include_assets=T)
```

### Heat Map

```{r}
dat = expand.grid(x = 1:5, y = 1:5)
dat = transform(dat, value = rnorm(25))
p5 = rPlot(x = 'bin(x, 1)', y = 'bin(y, 1)', color = 'value', data = dat, type = 'tile')
p5$print(include_assets=T)

```

### NBA Heat Map
```{r}
# NBA Heat Map
nba = read.csv('http://datasets.flowingdata.com/ppg2008.csv')
nba.m = ddply(melt(nba), .(variable), transform, rescale = rescale(value))
p6 = rPlot(Name ~ variable, color = 'rescale', data = nba.m, type = 'tile', height = 600)
p6$guides("{color: {scale: {type: gradient, lower: white, upper: steelblue}}}")
p6$print(include_assets=T)


```
### Another Facetted Scatter Chart

```{r}
#Another Facetted Scatter Chart
r1 = rPlot(mpg ~ wt | am + vs, data = mtcars, type = 'point',
  color = 'gear')
r1$print(include_assets=T)
```

## Morris

The next library we will be exploring is [MorrisJS](http://morrisjs.github.io/morris.js/), a javascript graph library based on the very competent `raphael.js`.

### Multi Line chart

```{r}
#Multi Line chart
data(economics, package = 'ggplot2')
econ = transform(economics, date = as.character(date))
m1 = mPlot(x = 'date', y = c('psavert', 'uempmed'), type = 'Line',
  data = econ)
m1$set(pointSize = 0, lineWidth = 1)
m1$print(include_assets=T)
```

### BSimple Bar Chart

```{r}
# BSimple Bar Chart
haireye = as.data.frame(HairEyeColor)
dat = subset(haireye, Sex == "Female" & Eye == "Blue")
p1 = mPlot(x = 'Hair', y = list('Freq'), data = dat, type = 'Bar', labels = list("Count"))
p1$set(hideHover = "auto")
p1$print(include_assets=T)

```

### Multi Bar Chart
```{r}
# Multi Bar Chart
dat = subset(haireye, Sex == "Female")
p2 = mPlot(Freq ~ Eye, group = "Hair", data = dat, type = "Bar", labels = 1:4)
p2$print(include_assets=T)

```

### Area Chart
```{r}
# Area Chart
data(economics, package = 'ggplot2')
dat = transform(economics, date = as.character(date))
p3 <- mPlot(x = "date", y = list("psavert", "uempmed"), data = dat, type = 'Line',
 pointSize = 0, lineWidth = 1)
p3$set(xLabelFormat = "#! function (x) { 
  return x.toString(); } 
!#")
p3$set(type = 'Area')
p3$print(include_assets=T)
```

## NVD3
[NVD3](http://nvd3.org/) is d3js library, which produces very nice interactive visualizations with little customization.

### Scatter Chart

```{r}

# Scatter Chart
p1 = nPlot(mpg ~ wt, group = 'cyl', data = mtcars, type = 'scatterChart')
p1$xAxis(axisLabel = 'Weight')
p1$print(include_assets=T)
```


### MultiBar Chart
```{r}
# MultiBar Chart
hair_eye = as.data.frame(HairEyeColor)
p2 = nPlot(Freq ~ Hair, group = 'Eye', data = subset(hair_eye, Sex == "Female"), type = 'multiBarChart')
p2$chart(color = c('brown', 'blue', '#594c26', 'green'))
p2$print(include_assets=T)
```


### MultiBar Horizontal Chart
```{r}
# MultiBar Horizontal Chart
p3 = nPlot(~ cyl, group = 'gear', data = mtcars, type = 'multiBarHorizontalChart')
p3$chart(showControls = F)
p3$print(include_assets=T)
```

### Pie Chart
```{r}
# Pie Chart
p4 = nPlot(~ cyl, data = mtcars, type = 'pieChart')
p4$print(include_assets=T)
```


###  Donut Chart
```{r}
###  Donut Chart
p5 = nPlot(~ cyl, data = mtcars, type = 'pieChart')
p5$chart(donut = TRUE)
p5$print(include_assets=T)
```

### Line Chart

```{r}
# Line Chart
data(economics, package = 'ggplot2')
p6 = nPlot(uempmed ~ date, data = economics, type = 'lineChart')
p6$print(include_assets=T)
```

### Line with Focus Chart

```{r}

ecm = reshape2::melt(economics[,c('date', 'uempmed', 'psavert')], id = 'date')
p7 = nPlot(value ~ date, group = 'variable', data = ecm, type = 'lineWithFocusChart')
p7$xAxis( tickFormat="#!function(d) {return d3.time.format('%b %Y')(new Date( d * 86400000 ));}!#" )
p7$print(include_assets=T)

```


### Stacked Area Chart
```{r}
# Stacked Area Chart
dat = data.frame(t=rep(0:23,each=4),var=rep(LETTERS[1:4],4),val=round(runif(4*24,0,50)))
p8 = nPlot(val ~ t, group =  'var', data = dat, type = 'stackedAreaChart', id = 'chart')
p8$print(include_assets=T)
```


### InteractiveGuidline(Multi-Tooltips) on Line

```{r}
# InteractiveGuidline(Multi-Tooltips) on Line

p9 = nPlot(value ~ date, group = 'variable', data = ecm, type = 'lineChart')
p9$xAxis( tickFormat="#!function(d) {return d3.time.format('%b %Y')(new Date( d * 86400000 ));}!#" )
#try new interactive guidelines feature
p9$chart(useInteractiveGuideline=TRUE)
p9$print(include_assets=T)

```


### showDistX and showDistY

```{r}
# showDistX and showDistY
p11 = nPlot(mpg ~ wt, group = 'cyl', data = mtcars, type = 'scatterChart')
p11$xAxis(axisLabel = 'Weight')
p11$chart(showDistX = TRUE, showDistY = TRUE)
p11$print(include_assets=T)
```


### Multi Chart

```{r}
# Multi Chart
p12 = nPlot(value ~ date, group = 'variable', data = ecm, type = 'multiChart')
p12$params$multi = list(
  uempmed = list(type="area",yAxis=1),
  psavert = list(type="line",yAxis=2)
)
p12$setTemplate(script = system.file("/libraries/nvd3/layouts/multiChart.html",package = "rCharts"))
p12$print(include_assets=T)
```


## xCharts

`xCharts` is a high quality looking charting library using `d3.js`, made by `TenXer`. Let's see it.


### Line Chart
```{r}
#xCharts
# Line Chart
require(reshape2)
uspexp = melt(USPersonalExpenditure)
names(uspexp)[1:2] = c('category', 'year')
x1 = xPlot(value ~ year, group = 'category', data = uspexp, 
  type = 'line-dotted')
x1$print(include_assets=T)

```




### Bar Chart

```{r}
# Bar Chart
haireye = subset(as.data.frame(HairEyeColor), Sex == "Male")
p2 <- xCharts$new()
p2$set(xScale = 'ordinal', yScale = 'linear', width = 600)
p2$layer(Freq ~ Hair, group = 'Eye', data = haireye, type = 'bar')
p2$print(include_assets=T)
```


## HighCharts

Highcharts is  also a  charting library written in pure JavaScript, offering an easy way of adding interactive charts to your web site or web application

### Scatter Chart

```{r}
# HighCharts
 # Scatter Chart
h1 = hPlot(Pulse ~ Height, data = MASS::survey, type = "scatter", group = "Exer")
h1$print(include_assets=T)
```

### Bubble Chart
```{r}
# Bubble Chart
h2 = hPlot(Pulse ~ Height, data = MASS::survey, type = "bubble", title = "Zoom demo", subtitle = "bubble chart", size = "Age", group = "Exer")
h2$chart(zoomType = "xy")
h2$exporting(enabled = F)
h2$print(include_assets=T)
```


### Pie Chart

```{r}
# Pie Chart
x = data.frame(key = c("a", "b", "c"), value = c(1, 2, 3))
h3 = hPlot(x = "key", y = "value", data = x, type = "pie")
h3$print(include_assets=T)
```

### Another Scatter Chart

```{r}
# Another Scatter Chart
h4 = hPlot(Pulse ~ Height, data = MASS::survey, type = 'scatter', group = 'Sex', radius = 6, group.na = "Not Available")
h4$colors('rgba(223, 83, 83, .5)', 'rgba(119, 152, 191, .5)', 'rgba(60, 179, 113, .5)')
h4$legend(align = 'right', verticalAlign = 'top', layout = 'vertical')
h4$plotOptions(scatter = list(marker = list(symbol = 'circle')))
h4$tooltip(formatter = "#! function() { return this.x + ', ' + this.y; } !#")
h4$print(include_assets=T)
```

### Multi Chart
```{r}
# Multi Chart
h5 = hPlot(freq ~ Exer, data = plyr::count(MASS::survey, c('Sex', 'Exer')), type = c('column', 'line'), group = 'Sex', radius = 6)
h5$print(include_assets=T)

```


```{r}
h6 = hPlot(x = "Wr.Hnd", y = "NW.Hnd", data = MASS::survey, 
  type = c("line", "bubble", "scatter"), group = "Clap", size = "Age")
h6$print(include_assets=T)
```

```{r}
# Bar Chart
a = hPlot(freq ~ Exer, data = plyr::count(MASS::survey, c('Sex', 'Exer')), type = 'bar', group = 'Sex', group.na = 'NA\'s')
a$plotOptions(bar = list(cursor = 'pointer', point = list(events = list(click = "#! function() { alert ('Category: '+ this.category +', value: '+ this.y); } !#"))))
a$print(include_assets=T)
```


```{r}
# Column Chart
df = data.frame(
    name = c("John", "John", "Joe", "Joe", "Jane", "Jane", "Janet", "Janet"),
    data = sample(1:10, 8, replace = T),
    stack = c(rep("male", 4), rep("female", 4))
)
a = rCharts:::Highcharts$new()
a$chart(type = "column")
a$plotOptions(column = list(stacking = "normal"))
a$xAxis(categories = c("Apples", "Oranges"))
invisible(sapply(unique(df$name), function(name){
    d = df[df$name == name, ]
    a$series(name = name, data = d$data, stack = d$stack[[1]])
}))
a$print(include_assets=T)
```

```{r}
# Leaflet
map3 = Leaflet$new()
map3$setView(c(41.7627, -72.6743), zoom = 13)
map3$marker(c(41.7627, -72.6743), bindPopup = "<p> Hi. I am a popup </p>")
map3$marker(c(41.70, -72.60), bindPopup = "<p> Hi. I am another popup </p>")
map3$print(include_assets=T)
```

## dygraphs for R
```{r}
library(dygraphs)
lungDeaths <- cbind(ldeaths, mdeaths, fdeaths)
dygraph(lungDeaths, main = "Deaths from Lung Disease (UK)") %>%
  dyOptions(colors = RColorBrewer::brewer.pal(3, "Set2"))
```



### Step Plots

By default dygraphs displays series as a line, you can however plot series as step chart as follows:

```{r}
lungDeaths <- cbind(mdeaths, fdeaths)
dygraph(lungDeaths, main = "Deaths from Lung Disease (UK)") %>%
  dyOptions(stepPlot = TRUE)
```

### Filling

You can also fill in the area underneath the series as well as customize the alpha value for the filling:
```{r}
dygraph(ldeaths, main = "Deaths from Lung Disease (UK)") %>%
  dyOptions(fillGraph = TRUE, fillAlpha = 0.4)

```

Point Display

You can include display of the individual points in a series as well as customize the size of the points:

```{r}
dygraph(ldeaths, main = "Deaths from Lung Disease (UK)") %>%
  dyOptions(drawPoints = TRUE, pointSize = 2)
```

Per-Series Options

All of the options above can also be set on a per-series basis using the dySeries function. For example:

```{r}
dygraph(lungDeaths, main = "Deaths from Lung Disease (UK)") %>%
  dySeries("mdeaths", drawPoints = TRUE, color = "blue") %>%
  dySeries("fdeaths", stepPlot = TRUE, fillGraph = TRUE, color = "red")
```

Note that you can specify colors globally (as demonstrated above) or per-series, but you cannot mix both styles of assigning colors.

Line Strokes

You can also customize the way that series lines are drawn. Here we draw a wider line with a custom stroke pattern (dashed line):

```{r}
dygraph(ldeaths, main = "Deaths from Lung Disease (UK)") %>%
  dySeries("V1", strokeWidth = 2, strokePattern = "dashed")
```

Range Selector

You can add range selector to the bottom of a dygraph that provides a straightforward interface for panning and zooming. For example:

```{r}
dygraph(nhtemp, main = "New Haven Temperatures") %>% 
  dyRangeSelector()
```
You can also specify an initial date range for the range selector:

```{r}
dygraph(nhtemp, main = "New Haven Temperatures") %>% 
  dyRangeSelector(dateWindow = c("1920-01-01", "1960-01-01"))
```

## References

* [rCharts](https://ramnathv.github.io/rCharts/)

* [dygraphs for R](https://rstudio.github.io/dygraphs/index.html)


