---
title: "Interpreting coefficients from interaction"
author: "Hicham Zmarrou"
date: "April 3, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Interaction between two categorical variables:

Let's make an hypothetical examples of a study, we measured the shoot length of some plant species under two different treatments: one is with increasing temperature (Low, High), the other is with three levels of nitrogen addition (A, B, C). We have made a completely factorial design and would like to look at the effect of these two treatments and their interactions on the shoot length.

```{r}
# interpreting interaction coefficients from lm first case two categorical
# variables
set.seed(12)
f1 <- gl(n = 2, k = 30, labels = c("Low", "High"))
f2 <- as.factor(rep(c("A", "B", "C"), times = 20))
modmat <- model.matrix(~f1 * f2, data.frame(f1 = f1, f2 = f2))
coeff <- c(1, 3, -2, -4, 1, -1.2)
y <- rnorm(n = 60, mean = modmat %*% coeff, sd = 0.1)
dat <- data.frame(y = y, f1 = f1, f2 = f2)
summary(lm(y ~ f1 * f2))
```
+ The first coefficient (0.97) is the intercept, so the shoot length for the Low temperature and the A nitrogen addition treatment. 
+ The second one (3) is the difference between the mean shoot length of the High temperature and the Low temperature treatment. 

+ Similarly the third and fourth one (-1.97, 4) are the mean shoot length difference between the treatment B-A and between the treatment C-A. 

+ The fifth and sixth one are more tricky, they are the added mean shoot length for pots with temperature High and nitrogen addition B or C as compared to the intercept. For example to get the mean shoot length for High temperature and nitrogen B we do: 0.97+3-1.97+0.98, this 0.98 is then the added difference for tese particular cases. Let's make some graph to show these coefficients:


## Interaction between two categorical variables:

Now let's turn to another case, there we are weighting standardize soil samples, we added a temperature treatment with two levels (Low, High) and we measured the soil nitrogen concentration, we would like to see the effects of the nitrogen concentration and its interaction with temperature on soil weight.

```{r}
# second case one categorical and one continuous variable
x <- runif(50, 0, 10)
f1 <- gl(n = 2, k = 25, labels = c("Low", "High"))
modmat <- model.matrix(~x * f1, data.frame(f1 = f1, x = x))
coeff <- c(1, 3, -2, 1.5)
y <- rnorm(n = 50, mean = modmat %*% coeff, sd = 0.5)
dat <- data.frame(y = y, f1 = f1, x = x)
summary(lm(y ~ x * f1))
```

This is an easy case, the first coefficient is the intercept, the second is the slope between the weight and the soil nitrogen concentration, the third one is the difference when the nitrogen concentration is 0 between the means for the two temperature treatments, and the fourth is the change in the slope weight~nitrogen between the Low and High temperature treatment.

```{r}
# plot (easy one)
ggplot(dat, aes(x = x, y = y, color = f1)) + geom_point() + geom_abline(xintercept = 1.17, 
    slope = 2.98, color = "red") + geom_abline(xintercept = 1.17 - 2.09, slope = 2.98 + 
    1.5, color = "blue") + geom_text(label = "y = 1.17 + 2.98 * x", x = 5, y = 10, 
    angle = 30, color = "black", size = 6) + geom_text(label = "y = (1.17 - 2.09) + (2.98 + 1.5) * x", 
    x = 5, y = 25, angle = 45, color = "black", size = 6)
```


## Interaction between two continuous variables

Now the last possible case could be something like a study where we measured the attack rates of carabids beetles on some prey and we collected two continuous variable: the number of prey item in the proximity of the beetles and the air temperature. We would like to see how these two variables influence the attack rates

```{r}
# third case interaction between two continuous variables
x1 <- runif(50, 0, 10)
x2 <- rnorm(50, 10, 3)
modmat <- model.matrix(~x1 * x2, data.frame(x1 = x1, x2 = x2))
coeff <- c(1, 2, -1, 1.5)
y <- rnorm(50, mean = modmat %*% coeff, sd = 0.5)
dat <- data.frame(y = y, x1 = x1, x2 = x2)
summary(lm(y ~ x1 * x2))
```


```{r}

# plot
ggplot(dat, aes(x = x2, y = y, color = x1)) + geom_point(size = 3) + scale_color_gradient(low = "red", 
    high = "green") + geom_abline(xintercept = 0.78, slope = -0.98, color = "red", 
    size = 1) + geom_abline(xintercept = 0.78, slope = 15, color = "green", 
    size = 1) + geom_abline(xintercept = 0.78, slope = (5 * 1.5) - 1, color = "brown", 
    size = 1) + geom_text(label = "y = 0.78 + (-0.98+1.49*0)*x2", x = 11, y = 0, 
    color = "red", size = 5) + geom_text(label = "y = 0.78 + (-0.98+1.49*5)*x2", 
    x = 10, y = 80, color = "brown", angle = 15, size = 5) + geom_text(label = "y = 0.78 + (-0.98+1.49*15)*x2", 
    x = 9, y = 160, color = "green", angle = 30, size = 5) + ylim(c(-20, 250))
```

